---
title: "ATS Study"
author: "Supanat Thitipatarakorn"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is baseline (Phase 1) data from the ATS study including only MSM and excluding all TGW and 1 MSM who did not complete baseline questionnaire.

```{r}
# Install packages
library("dplyr")
library("forcats")
library("gtsummary")
library(nnet)
library(car)
```

## Data Exploration

```{r include=FALSE}
# Read in the data
df <- read.csv("data_ATS_stageI_26082024.csv", na.string = "")

# dimension of the data frame
dim(df) # There are 1394 observations and 141 variables in the main file

# check duplicate
n_distinct(df$patientid) # no duplicated patient IDs

# column names
colnames(df)

# filter out unused patient IDs
df2 <- df %>%
  filter(ats_6m != "[2]Skip/prefer not to answer") # exclude 19 participants who did not answer questions about ATS, so 1375 participants left

# fill missing values
df3 <- df2 %>%
   mutate(income_4gr_fill = case_when(
     income_2gr == "[0]No income" ~ "[1]<=15,000", #group no income into the ≤15,000 group
     .default = as.character(income_4gr)),
     current_smoker = case_when(
       is.na(q28_1) ~ q28,
       .default = as.character(q28_1)
     ),
     income_4gr_fill = fct_na_value_to_level(income_4gr, level = "Unknown"),
     syphilis_6m_crf_fill = fct_na_value_to_level(syphilis_6m_crf, level = "No"),
     gonorrhea_6m_crf_fill = fct_na_value_to_level(gonorrhea_6m_crf, level = "No"),
     ct_6m_crf_fill = fct_na_value_to_level(ct_6m_crf, level = "No"),
     any_bact_sti = case_when(
       syphilis_6m_crf == "[1]Syphilis" ~ "Yes",
       gonorrhea_6m_crf == "[1]Gonorrhea" ~ "Yes",
       ct_6m_crf == "[1]Chlamydia" ~ "Yes",
       .default = "No"
     )) 

# mutate other variables
df4 <- df3 %>%
  mutate(antihiv_noincon = ifelse(antihiv == "[3]Inconclusive", NA, antihiv), # hide anti-HIV inconclusive
         lifetime_partners = cut(q11, breaks = c(1, 2, 6, 11, Inf), include.lowest = TRUE, right = FALSE), # number of lifetime partners
         lifetime_partners_2gr = cut(q11, breaks = c(1, 6, Inf), include.lowest = TRUE, right = FALSE),
         income_15000 = case_when(
           income_4gr_fill == "[1]<=15,000" ~ "<=15,000",
           income_4gr_fill == "[2]15,001-20,000" ~ ">15,000",
           income_4gr_fill == "[3]20,001-30,000" ~ ">15,000",
           income_4gr_fill == "[4]>30000" ~ ">15,000"
         ),
         income_30000 = case_when(
           income_4gr_fill == "[1]<=15,000" ~ "<=30,000",
           income_4gr_fill == "[2]15,001-20,000" ~ "<=30,000",
           income_4gr_fill == "[3]20,001-30,000" ~ "<=30,000",
           income_4gr_fill == "[4]>30000" ~ ">30,000"
         ),
         education = case_when(
           q1 == "[1]Primary School" ~ "Lower than university",
           q1 == "[2]Secondary School" ~ "Lower than university",
           q1 == "[3]Technical School or Diploma" ~ "Lower than university",
           q1 == "[4]University or higher" ~ "University or higher",
           q1 == "[6]Skip/prefer not to say" ~ NA
         ),
         q27 = as.factor(q27)
         )

# save to final DF
df_final <- df4
```

`q1` = education
`q11` = No. of lifetime sex partners
`q23` = group sex
`q27` = current PrEP user
`q29_2gr` = alcohol
`q29_auditc`= alcohol AUDIT-C

## Summary table
```{r summary_table}
# summary table
df_tbl_summary <- df_final %>%
    select(ats_6m_2gr, age, age_4gr, status_crf, income_4gr_fill, income_15000, income_30000, education, syphilis_6m_crf_fill, gonorrhea_6m_crf_fill, ct_6m_crf_fill, any_bact_sti, antihiv, antihiv_noincon, q1, lifetime_partners, lifetime_partners_2gr, q23, q27, current_smoker, q29_2gr, q29_auditc, ptsd_2gr, hads_depress_3gr, hads_anxiety_3gr) %>%      # select variable of interest 
  mutate(income_4gr_fill = factor(income_4gr_fill) %>% forcats::fct_explicit_na(),
         lifetime_partners = factor(lifetime_partners) %>% forcats::fct_explicit_na(),
         lifetime_partners_2gr = factor(lifetime_partners_2gr) %>% forcats::fct_explicit_na(),
         q23 = factor(q23) %>% forcats::fct_explicit_na(),
         q27 = factor(q27) %>% forcats::fct_explicit_na(),
         current_smoker = factor(current_smoker) %>% forcats::fct_explicit_na(),
         q29_2gr = factor(q29_2gr) %>% forcats::fct_explicit_na(),
         q29_auditc = factor(q29_auditc) %>% forcats::fct_explicit_na(),
         income_15000 = factor(income_15000) %>% forcats::fct_explicit_na(),
         income_30000 = factor(income_30000) %>% forcats::fct_explicit_na(),
         education = factor(education) %>% forcats::fct_explicit_na(),
         ptsd_2gr = factor(ptsd_2gr) %>% forcats::fct_explicit_na(),
         hads_depress_3gr = factor(hads_depress_3gr) %>% forcats::fct_explicit_na(),
         hads_anxiety_3gr = factor(hads_anxiety_3gr) %>% forcats::fct_explicit_na()
         )

# by = hads_depress_3gr
df_tbl_summary %>% tbl_summary(                                                
        statistic = list(all_continuous() ~ c("{N_nonmiss}",
                                              "{median} ({p25} to {p75})", 
                                              "{min} to {max}"),        # stats and format for continuous variables
                         all_categorical() ~ "{n} ({p}%)"),          # stats and format for categorical variables
        type = list(age ~ "continuous2",
                    all_dichotomous() ~ "categorical"),             # force show all levels
        digits = list(all_continuous() ~ 1,                         # rounding for continuous columns
                      all_categorical() ~ c(0,1)),                       # rounding for categorical columns
        label = list(age ~ "Age (years)",
                     age_4gr ~ "Age (years)"),
        by = hads_depress_3gr,
        missing = "ifany",
        percent = "row"
    ) %>%
  add_n %>% 
  # add_p(pvalue_fun = ~style_pvalue(.x, digits = 3)) %>% 
  add_overall() %>%
  modify_caption("**HADS-D**")

# by = hads_anxiety_3gr
df_tbl_summary %>% tbl_summary(                                                
        statistic = list(all_continuous() ~ c("{N_nonmiss}",
                                              "{median} ({p25} to {p75})", 
                                              "{min} to {max}"),        # stats and format for continuous variables
                         all_categorical() ~ "{n} ({p}%)"),          # stats and format for categorical variables
        type = list(age ~ "continuous2",
                    all_dichotomous() ~ "categorical"),             # force show all levels
        digits = list(all_continuous() ~ 1,                         # rounding for continuous columns
                      all_categorical() ~ c(0,1)),                       # rounding for categorical columns
        label = list(age ~ "Age (years)",
                     age_4gr ~ "Age (years)"),
        by = hads_anxiety_3gr,
        missing = "ifany",
        percent = "row"
    ) %>%
  add_n %>%
  add_overall() %>%
  modify_caption("**HADS-A**")

# by = ptsd_2gr
df_tbl_summary %>% tbl_summary(                                                
        statistic = list(all_continuous() ~ c("{N_nonmiss}",
                                              "{median} ({p25} to {p75})", 
                                              "{min} to {max}"),        # stats and format for continuous variables
                         all_categorical() ~ "{n} ({p}%)"),          # stats and format for categorical variables
        type = list(age ~ "continuous2",
                    all_dichotomous() ~ "categorical"),             # force show all levels
        digits = list(all_continuous() ~ 1,                         # rounding for continuous columns
                      all_categorical() ~ c(0,1)),                       # rounding for categorical columns
        label = list(age ~ "Age (years)",
                     age_4gr ~ "Age (years)"),
        by = ptsd_2gr,
        missing = "ifany",
        percent = "row"
    ) %>%
  add_n %>%
  add_overall() %>%
  modify_caption("**PTSD**")

```

---

## Multinomial logistic regression model
## HADS-D as an outcome
### Univariate
```{r}
# select variables for the models
df_model <- df_final %>%
  select(ats_6m_2gr, age_4gr, income_15000, income_30000, education, any_bact_sti, antihiv_noincon, q23, q27, current_smoker, q29_2gr, q29_auditc, ptsd_2gr, hads_anxiety_3gr, hads_depress_3gr) %>%
  mutate(ptsd_abnormal = case_when(
    ptsd_2gr == "[0]Normal score 0-2" ~ 0,
    ptsd_2gr == "[1]Abnormal score 3-4" ~ 1
  )) %>%
  select(-ptsd_2gr)

# univariate models
univariate_hasd_d <- df_model %>%
    tbl_uvregression(
        method = multinom,
        y = hads_depress_3gr,
        method.args = list(family = binomial(link = "logit")),
        exponentiate = TRUE,
        pvalue_fun = ~style_pvalue(.x, digits = 3)
    ) %>%
  bold_p()
univariate_hasd_d
```

### Multivariate

Include variables with p<0.05 from univariate analysis in the multivariate model.

```{r}
# multivariate model 1
m_hasd_d_1 <- multinom(hads_depress_3gr ~ ats_6m_2gr + income_15000 + current_smoker + ptsd_abnormal + hads_anxiety_3gr + q29_auditc, family = binomial(link = "logit"), data = df_model)

m_hasd_d_1 %>%
  tbl_regression(
    exponentiate = TRUE, 
    pvalue_fun = ~style_pvalue(.x, digits = 3),
  ) %>% 
  bold_labels() %>%
  bold_p() %>%
  add_glance_source_note(
    include = c(AIC)
  )
```

```{r}
# Calculate GVIF separately for 2 logistic regression models because GVIF isn't designed for the multinomial model from nnet.
df_model_vif <- df_model %>%
  mutate(hads_depress_borderline = case_when(
    hads_depress_3gr == "[0]Normal score 0-7" ~ 0,
    hads_depress_3gr == "[1]Borderline score 8-10" ~ 1,
    hads_depress_3gr == "[2]Abnormal score 11-21" ~ NA
  ),
  hads_depress_abnormal = case_when(
    hads_depress_3gr == "[0]Normal score 0-7" ~ 0,
    hads_depress_3gr == "[1]Borderline score 8-10" ~ NA,
    hads_depress_3gr == "[2]Abnormal score 11-21" ~ 1
  ))

vif(glm(hads_depress_borderline ~ income_15000 + current_smoker + ptsd_abnormal + hads_anxiety_3gr + q29_auditc, data=df_model_vif,family="binomial"))
vif(glm(hads_depress_abnormal ~ income_15000 + current_smoker + ptsd_abnormal + hads_anxiety_3gr + q29_auditc, data=df_model_vif,family="binomial"))
# The high adjusted GVIF values (similar to VIF) i.e. > 2 indicate the presence of moderate to strong multicollinearity. 
# None of the adjusted GVIF is > 2 in these models.
```

Remove variables with p>0.05 from model 1

```{r}
# multivariate model 2 (including varoables with p<0.05 from model 1)
m_hasd_d_2 <- multinom(hads_depress_3gr ~ ats_6m_2gr + income_15000 + current_smoker + ptsd_abnormal + hads_anxiety_3gr, family = binomial(link = "logit"), data = df_model)

m_hasd_d_2 %>%
  tbl_regression(
    exponentiate = TRUE, 
    pvalue_fun = ~style_pvalue(.x, digits = 3),
  ) %>% 
  bold_labels() %>%
  bold_p() %>%
  add_glance_source_note(
    include = c(AIC)
  )
```

The AIC became higher, so model 1 is preferred.

---

## HADS-A as an outcome
### Univariate
```{r}
# univariate models
univariate_hasd_a <- df_model %>%
    tbl_uvregression(
        method = multinom,
        y = hads_anxiety_3gr,
        method.args = list(family = binomial(link = "logit")),
        exponentiate = TRUE,
        pvalue_fun = ~style_pvalue(.x, digits = 3)
    ) %>%
  bold_p()
univariate_hasd_a
```

### Multivariate
```{r}
# model 1
m_hasd_a_1 <- multinom(hads_anxiety_3gr ~ ats_6m_2gr + age_4gr + income_15000 + education + antihiv_noincon + current_smoker + ptsd_abnormal + hads_depress_3gr, family = binomial(link = "logit"), data = df_model)
# can't include q27 (PrEP user) in the model because of a single-level issue.

m_hasd_a_1 %>%
  tbl_regression(
    exponentiate = TRUE, 
    pvalue_fun = ~style_pvalue(.x, digits = 3),
  ) %>% 
  bold_labels() %>%
  bold_p() %>%
  add_glance_source_note(
    include = c(AIC)
  )
```

```{r}
# Calculate GVIF separately for 2 logistic regression models because GVIF isn't designed for the multinomial model from nnet.
df_model_vif <- df_model %>%
    mutate(hads_anxiety_borderline = case_when(
        hads_anxiety_3gr == "[0]Normal score 0-7" ~ 0,
        hads_anxiety_3gr == "[1]Borderline score 8-10" ~ 1,
        hads_anxiety_3gr == "[2]Abnormal score 11-21" ~ NA
    ),
    hads_anxiety_abnormal = case_when(
        hads_anxiety_3gr == "[0]Normal score 0-7" ~ 0,
        hads_anxiety_3gr == "[1]Borderline score 8-10" ~ NA,
        hads_anxiety_3gr == "[2]Abnormal score 11-21" ~ 1
    ))

vif(glm(hads_anxiety_borderline ~ ats_6m_2gr + age_4gr + income_15000 + education + antihiv_noincon + current_smoker + ptsd_abnormal + hads_depress_3gr, data=df_model_vif,family="binomial"))
vif(glm(hads_anxiety_abnormal ~ ats_6m_2gr + age_4gr + income_15000 + education + antihiv_noincon + current_smoker + ptsd_abnormal + hads_depress_3gr, data=df_model_vif,family="binomial"))
# The high adjusted GVIF values (similar to VIF) i.e. > 2 indicate the presence of moderate to strong multicollinearity. 
# None of the adjusted GVIF is > 2 in these models.
```
Next, include only variables with p <0.05 from model 1 into model 2.

```{r}
# model 2
m_hasd_a_2 <- multinom(hads_anxiety_3gr ~ ats_6m_2gr + age_4gr + income_15000 + antihiv_noincon + ptsd_abnormal + hads_depress_3gr, family = binomial(link = "logit"), data = df_model)

m_hasd_a_2 %>%
  tbl_regression(
    exponentiate = TRUE, 
    pvalue_fun = ~style_pvalue(.x, digits = 3),
  ) %>% 
  bold_labels() %>%
  bold_p() %>%
  add_glance_source_note(
    include = c(AIC)
  )
```

The AIC of model 2 is larger than that of model 1, so model 1 is preferred.

---

## PTSD as an outcome
### Univariate
```{r}
# univariate models
univariate_ptsd <- df_model %>%
    tbl_uvregression(
        method = glm,
        y = ptsd_abnormal,
        method.args = list(family = binomial(link = "logit")),
        exponentiate = TRUE,
        pvalue_fun = ~style_pvalue(.x, digits = 3)
    ) %>%
  bold_labels() %>%
  bold_p()
univariate_ptsd
```

### Multivariate

Model 1: include variables with p<0.05 from univariate analysis.

```{r}
# model 1
m_ptsd_1 <- glm(ptsd_abnormal ~ ats_6m_2gr + education + q27 + current_smoker + q29_auditc + hads_anxiety_3gr + hads_depress_3gr, family = binomial(link = "logit"), data = df_model)

multivariate_ptsd_1 <- m_ptsd_1 %>%
  tbl_regression(
    exponentiate = TRUE, 
    pvalue_fun = ~style_pvalue(.x, digits = 3),
  ) %>% 
  bold_labels() %>%
  bold_p() %>%
  add_glance_table(
    include = c(AIC)
  )

# model 2 (selecting variables with p<0.05 from model 1)
m_ptsd_2 <- glm(ptsd_abnormal ~ ats_6m_2gr + q29_auditc + hads_anxiety_3gr + hads_depress_3gr, family = binomial(link = "logit"), data = df_model)

multivariate_ptsd_2 <- m_ptsd_2 %>%
  tbl_regression(
    exponentiate = TRUE, 
    pvalue_fun = ~style_pvalue(.x, digits = 3),
  ) %>% 
  bold_labels() %>%
  bold_p() %>%
  add_glance_table(
    include = c(AIC)
  )

tbl_merge(list(univariate_ptsd, multivariate_ptsd_1, multivariate_ptsd_2),
          tab_spanner = c("**Univariable**", "**Multivariable 1**", "**Multivariable 2**"))

# VIF
vif(m_ptsd_1)
# The high adjusted GVIF values (similar to VIF) i.e. > 2 indicate the presence of moderate to strong multicollinearity. 
# None of the adjusted GVIF is > 2 in this model.
```

The larger model (model 1) has a smaller AIC value. Hence, model 1 is preferred.